---
title: "genetic_structure"
author: "Thomas Francisco"
date: "2024-01-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pcadapt)
library(ggplot2)
library(LEA)
library(rnaturalearth)
library(scatterpie)
```

The goal here is to analyse the genetic structure of the species and check if the structure of some individuals/populations is not drive by the Na. To do so, we will plot the size of the sampled points on the PCa depending on their Na percentage.
```{r PCa with Na}

#file with Na
data_i= "C:/Users/tfrancisco/Documents/stage M2/2024_new_filtration/data.Rdata"
load(file = data_i)

#We're gonna used the Pcadapt function to perform the PCa. The format of the input dataset is: row: SNPs, columns: individuals

#If necessary, transpose the dataset
data_PCa <- data.frame(t(data))

#format for pcadapt
data_pcadapt <- read.pcadapt(data)

#perform the Pca
Pca <-pcadapt(data_pcadapt,K=10,method = "mahalanobis")

#
plot(Pca$singular.values, type="b")


score_Pca <- data.frame(Pca$scores[,1:3]);row.names(data)=row.names(data)

na_percentage <- rowMeans(is.na(data_PCa))*12.5

ggplot_representation <- ggplot(data=score_Pca,aes(x=X1,y=X2))+
  geom_point(size=na_percentage)

ggplot_representation


```

# Ancestry coefficients

Here we are going to use the LEA package to estimate the ancestry of the individuals. We are going to produce a map of the points and for each of them we will have a pie with his ancestry coefficients. 
First, to use the LEA package, we need to transform the genetic dataset into the geno format (SNPs in rows, individuals in columns and 0,1,2, 9(for Nas)). 
```{r LEA genetic structure}
#replace the Na per 0 
data_PCa_geno <- data.frame(replace(data_PCa, is.na(data_PCa),9)) 
#write the file in geno format: 
write.geno(data_PCa_geno,"C:/Users/tfrancisco/Documents/stage M2/2024_new_filtration/data_PCa_geno.geno")#here the input of write.geno is a dataset with individuals in rows, snps in columns and Na in 9

#perform the ancestry calculation
project_1 <- snmf("C:/Users/tfrancisco/Documents/stage M2/2024_new_filtration/data_PCa_geno.geno",K=1:10,entropy=T,repetitions=1,project="new")
plot(project_1, col = "blue", pch = 19, cex = 1.2)

#plot the results
best = which.min(cross.entropy(project_1, K = 6))
my.colors <- c("tomato", "lightblue",
"olivedrab", "gold")
barchart(project_1, K = 6, run = best,
border = NA, space = 0,
col = my.colors,
xlab = "Individuals",
ylab = "Ancestry proportions",
main = "Ancestry matrix") -> bp
axis(1, at = 1:length(bp$order),
labels = bp$order, las=1,
cex.axis = .4)

#we can also output a matrix of the ancestry coeff for each individuals
qmatrix = LEA::Q(project_1, K = 6)
qmatrix_ID=data.frame(row.names(data_PCa_geno),qmatrix);colnames(qmatrix_ID)=c("ID","Group1","Group2","Group3","Group4","Group5","Group6")


#The next step will be to create the map. For this, we first need to mean the values at the populations levels, add the coordinates and then plot this on a map.

meta_data_vcf=read.csv("C:/Users/tfrancisco/Documents/stage M2/Worldclim/293_indiv.csv",h=T,sep=";",dec=",")
data_tot_maps <- merge(meta_data_vcf[,c("ID","demes","Long..WGS84.","Lat..WGS84.")],qmatrix_ID,"ID");colnames(data_tot_maps)=c("ID", "Pop","Longitude","Latitude","Group1","Group2","Group3","Group4","Group5","Group6");data_tot_maps$Longitude=as.numeric(data_tot_maps$Longitude);data_tot_maps$Latitude=as.numeric(data_tot_maps$Latitude)

#Map

# Obtain the blank map of the region
admin <- ne_countries(scale = "medium", returnclass = "sf")

# Plot the map with scatterpie
map_pop <- ggplot() + 
  geom_sf(data = admin, fill = gray(0.92), size = 0) +#background of the map
  geom_scatterpie(
    data = data_tot_maps,
    aes(x = Longitude, y = Latitude),
    cols = c("Group1", "Group2", "Group3","Group4","Group5","Group6"), # Adjust columns accordingly
    size = 0,#size of the bold (=gras)
    alpha = 1 #transparence
  ) +
  coord_sf(xlim = c(-2, 10), ylim = c(42, 48), expand = FALSE) +#extension of the map
  theme_minimal() +
  labs(title = "Genetic Structure for each Populations")

print(map_pop)



```


